Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-10-13 paper.tex
updated captions+sec.VB

% --- journals --- %
\newcommand{\aap}{A{\&}A}
\newcommand{\aaps}{A{\&}AS}
\newcommand{\aj}{AJ}
%\newcommand{\apj}{ApJ}
\newcommand{\apjl}{ApJL}
\newcommand{\apjs}{ApJS}
\newcommand{\apss}{Ap{\&}SS}
\newcommand{\araa}{ARA{\&}A}
%\newcommand{\prd}{Phys. Rev. D}
%\newcommand{\pre}{PRE}
\newcommand{\mnras}{MNRAS}
\newcommand{\ssr}{SSRv}
%\newcommand{\nat}{Nature}
\newcommand{\physrep}{Phys.~Rep.}
\newcommand{\jcomp}{J.~Comput.~Phys.}
\newcommand{\jcap}{JCAP}

\newcommand{\Fermi}{{\em Fermi}\xspace}

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
\newcommand{\egt}{\tilde{E}_\gamma}
\newcommand{\eet}{\tilde{E}_\rmn{e}}
\newcommand{\bra}{\langle}
\newcommand{\ket}{\rangle}
\newcommand{\vst}{\vspace{-0.14mm}}
\newcommand{\vstt}{\vspace{-0.0mm}}
\newcommand{\el}{\rmn{el}}
\newcommand{\mytimes}{\hspace{-0.5mm}\times\hspace{-0.5mm}}
\newcommand{\hst}{\hspace{-1.0mm}}

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). Furthermore,
since most of the gamma-ray flux is expected to come from dense
subhalos within clusters which have a velocity dispersion close to the
velocity limit $v_\rmn{sat}$ where the SFE saturates, we use
$S=\B_\sfe(v_\rmn{sat})\approx 530$ for the cluster
halos.\footnote{Note that $S$ is strictly speaking the Sommerfeld
  enhanced annihilation cross-section, but here we have additionally
  included the order unity constraint that ensures that the relic
  density is not over-depleted at freeze-out.} However, for figures
where we neglect the substructure boost, we adopt a velocity dependent
SFE that is normalized to fit the electron and positron excess observed at
Earth\del{observed by the PAMELA and \Fermi satellites},
\begin{equation}
\B_\sfe(\sigv) = \B_\rmn{sfe,MW} \frac{v_\rmn{MW}}{\sigv} =
70 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
\C{where the local boost factor required to explain the data is
  $\B_\rmn{sfe,MW}=300$.} \del{Here the boost factor needed to explain
  the PAMELA/\Fermi excess of electrons and positrons
  $\B_\rmn{sfe,MW}=300$,} The velocity dispersion for the MW,
$v_\rmn{MW} \approx 220\,\rmn{km\,s}^{-1}$, and the \C{mass dependent}
velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. \del{Note that, clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/fit.ds.flux.bw.eps}
 \includegraphics[width=0.49\columnwidth]{figures/fit.epflux.int.bw.eps}
 \caption{\it (Color online) Source functions for different dark
   matter (DM) models. We show the simulated data from {\sc DarkSUSY}
   \cite{ds}, and that generated by
   \protect\cite{2011JCAP...03..019C,2011JCAP...03..051C}; the solid
   lines show the fit to the data. LEFT PANEL: normalized differential
   continuum spectra for four different DM benchmark (BM) models;
   $\Ip$ model (red circles), $\Jp$ (orange squares), $\Kp$ (green
   diamonds), and $\Js$ (blue triangles). We use Eq.~(\ref{eq:bm_cont})
   to fit the continuum spectra. RIGHT PANEL: number of electron and
   positron per DM annihilation above the electron energy $\ee$ for
   different DM models; $\Ip$ BM model (red circles), $\Jp$ BM model
   (orange squares), $\Kp$ BM model (green diamonds), and $\Js$ BM
   model (blue triangles), leptophilic (LP) DM annihilating indirectly
   into electrons and positrons (purple $+$) and into muons (black
   $\times$) without considering Sommerfeld boosts. We use
   Eq.~(\ref{eq:bm_elec}) and Eqs.~(\ref{eq:lp_mu}-\ref{eq:lp_ep}) to
   fit the spectra of electrons and positrons from BM and LP models,
   respectively.}
 \label{fig:q_DM}
\end{minipage}
\end{figure*}

@@

by $\qCR(\eg,r)$ (see Sect.~\ref{sect:CRs} and

@@

(see Eq.~\ref{eq:B_sfe} and Sect.~\ref{sect:LP}) and from substructure

@@

Eqs.~\ref{eq:rho_sub}-\ref{eq:xvir} and Sect.~\ref{sect:subst}). The

@@

We now derive the squared density profile for the substructures
using
\begin{eqnarray}
\rho_\rmn{sub}^2(r) &=& \frac{\dd L_\rmn{sub}}{\dd V} \frac{1}{A_{\rmn{dm}}}\,,\label{eq:rho_sub}\\
L_{200\sm} &=& A_{\rmn{dm}} \int_{\rvir} \dd r' 4\pi r'^2 \rho^2(r')\,,\label{eq:Lsm}\\
 \frac{\dd L_\rmn{sub}}{\dd V} &=& a_0\,a_1\,C(\mvir)\,x^{g(x)}\,
\left[\frac{1+a_2\log(x)}{4\pi\,r^3}\right]\,,\nonumber\\
\\
x &=& r/\rvir\,,\quad g(x) = a_2+a_1\,x^{a_2}\,.\label{eq:xvir}
\end{eqnarray}
Here, $A_{\rmn{dm}} = q_{\rmn{sm},i}(\eg,R) / \rho(R)^2$ represents
the particle physics factor, where $q_{\rmn{sm},i}(\eg,R)$ is defined
by Eq.~(\ref{eq:q_sm}). The different density profiles have some
impact on the luminosity from annihilating DM, although the details of
the density profile can be neglected compared to the dominating boost
from substructures (assuming DM to be cold). In
Fig.~\ref{fig:radial_lum} we compare the radial dependence of the
accumulative luminosity from different smooth cluster density profiles
to the boosted luminosity due to substructure for different mass
scales. We recalculate the overdensity, $\rhos$, for the shallower
density profile with $\beta=0.6$ and rescale the concentration
parameter in Eq.~(\ref{eq:cfit}) with $300/160$
\cite{2011ApJ...728L..39N} to account for the more centralized scale
radius in cluster with a shallow inner slope. The emission of this
profile with $\beta=0.6$ is about 30\% larger within $\rvir$ compared
to a cuspy NFW ($\beta=1.0$). This difference is built up within
$0.1\rvir$ (i.e. close to $r_\s$). Hence the slope of the central part
of a cluster has little influence for the DM luminosity within $\rvir$
as long as the degeneracy in the lensing mass measurements with $r_\s$
(which decreases for decreasing inner slope) has been taken into
account. The emission from an Einasto density profile is about 50\%
larger than the cuspy NFW profile in the periphery of the cluster,
where the difference is mainly built up at a few percent of
$\rvir$. The increase in luminosity due to substructures is negligible
in the center of halos, but integrated out to $\rvir$ the boost
relative to the smooth emission profile amounts to approximately 20
for dwarf galaxies, $200$ for galaxies, and $10^3$ for galaxy
clusters. \C{We stress that these boost factors are only realized in
  the region around the virial radius of each respective halo which
  is mostly tidally stripped for dwarfs in the MW. Hence a more
  realistic boost from substructures is probably much smaller in
  satellite dwarf galaxies. In addition,} these boost factors are only
realized for direct annihilation emission (continuum emission or final
state radiation) or IC scattering off homogeneous seed photon fields
(CMB). For IC scattering of SD photons, the overlap of final state
leptons and SD photons is smaller which causes the substructure boost
over the smooth emission to be reduced by roughly two orders of
magnitudes. It should also be noted that the substructure boosted
fluxes from clusters are much more extended than for dwarf galaxies,
hence more difficult to detect with Cherenkov telescopes. In fact, it
was shown in Sanchez-Conde et al. \cite{2011arXiv1104.3530S} that
inside $0.1$~deg, where the sensitivity of Cherenkov telescopes is
maximized, the expected DM flux including substructures from the
brightest dwarf galaxy is about an order magnitude higher than the
brightest cluster. However, that work assumed a substructure boost of
about 50 in the most massive cluster halos which is more than an order
magnitude smaller than what we use in this work, hence in projection
we expect this difference to be a factor few smaller.

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/dens.prof.bw.eps}
 \caption{\it (Color online) Radial dependence of DM annihilation
   luminosity of smooth halo and substructures. The solid lines show
   the accumulative smooth luminosity from a cluster with the mass
   $\mvir=10^{15}\,\msun$ for three different density profiles; an
   Einasto profile with $\alpha=0.17$ (light blue), a cuspy NFW
   profile with $\beta=1.0$ (thick dark blue), and a core NFW profile
   with $\beta=0.6$ (thin purple). The dashed lines show the
   accumulative luminosity from substructures for three different mass
   scales; an $\mvir=10^{15}\,\msun$ galaxy cluster (thick red), an
   $\mvir=10^{12}\,\msun$ galaxy (orange), and an
   $\mvir=10^{8}\,\msun$ dwarf galaxy (thin green). All luminosities
   have been normalized with the luminosity within $\rvir$ from a
   cuspy NFW profile. We have assumed the standard value for the
   limiting substructure mass of $M_\rmn{lim}=10^{-6}\,\msun$. Note
   the large expected boost from substructures in clusters
   ($\sim1000$), and the relative small boost in dwarf galaxies
   ($\sim20$).}
 \label{fig:radial_lum}
\end{figure}

In Fig.~\ref{fig:radial_emis} we show the radial regions that dominate
the DM annihilation luminosity, in particular we show the differential
contribution to the DM luminosity per logarithmic interval in radius
for three different mass scales. Solving for the maximum of this
curve, $\dd^2L_{200\sm}/\dd (\log x)^2=0$ in combination with
Eq.~(\ref{eq:Lsm}), we find that the luminosity from the smooth NFW
profile peaks at $r \simeq \rvir/3c$. Despite the cuspy nature of the
density profile, the luminosity is not dominated by the central region
but by the transition region where the profile steepens because of the
larger volume available there. For large clusters with typical
concentrations of $c = \rvir/r_\s \simeq 4$, the luminosity from the
smooth profile is focused to the regime around 10\% of $\rvir$. In
contrast, emission from substructures is mainly contributed by the
outer parts of DM halos. As shown in Fig.~\ref{fig:radial_emis}, the
product of annihilation emissivity and emission volume increases
towards $\rvir$ and only starts to drop outside this radius. Note
that, even though most of the substructure mass density has been
erased in the central regions of DM halos, a cluster in projection has
a significant enhancement due to substructures at a radius of just a
few percent of $\rvir$.
\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/emissiv.sub.bw.eps}
  \caption{\it (Color online) Plot that shows where most of the DM
    annihilation luminosity originates. We show the differential
    contribution to the DM annihilation luminosity per logarithmic
    interval in radius which corresponds to $x^3\rho^2/\rho^2_{200}$
    where $\rho_{200}$ is the density at $\rvir$. The solid lines
    represent cuspy $\beta=1.0$ NFW density profiles for three
    different mass scales; an $\mvir=10^{15}\,\msun$ galaxy cluster
    (thick red), an $\mvir=10^{12}\,\msun$ galaxy (orange), and an
    $\mvir=10^{8}\,\msun$ dwarf galaxy (thin green). The dashed lines
    show the contribution from substructures for the same three mass
    scales. All luminosities have been normalized with the luminosity
    within $\rvir$ from a cuspy NFW profile. We have assumed the
    standard value for the limiting substructure mass of
    $M_\rmn{lim}=10^{-6}\,\msun$. For the smooth profile, the majority
    of the flux is delivered by a region around $r_\s/3$ as indicated
    by the maximum value of the curves. In contrast, for substructures
    the emission is dominated by regions around $r_{200}$.}
  \label{fig:radial_emis}
\end{figure}

@@

The electrons injected from annihilating DM also suffer from diffusive
and radiative losses. Hence we have to calculate the equilibrium
spectrum of the electrons plus positrons denoted by $\dd n_\e/\dd \ee$
in Eq.~(\ref{eq:ICemiss}). We derive this stationary solution using
the cosmic ray transport equation (neglecting convection and
re-acceleration effects):
\begin{eqnarray}
\label{eq:CRdiffusion}
\frac{\partial}{\partial t}\left(\frac{\dd n_\e}{\dd \ee}\right) =
&\nabla& \left[D\left(\ee,\bx\right)\nabla\frac{\dd n_\e}{\dd \ee}\right] +
\frac{\partial}{\partial \ee}
\left[b\left(\ee,\bx\right) \frac{\dd n_\e}{\dd \ee}\right]
 \nonumber \\
&+& q_\e(\ee,\bx)\,,
\end{eqnarray}
where $D(\ee,\bx)$ denotes the diffusion coefficient and $b(\ee,\bx)$
the energy loss term. The source function $q_\e(\ee,\bx)$ yields the
number of electrons and positrons produced per unit time, energy and
volume element at the position $\bx$:
\begin{equation}
q_\e(\ee,r)=\sum_f\frac{\dd N_\e^f}{\dd \ee}(\ee) B_f \Gamma_f(r) \,,
\end{equation}
where the annihilation rate density $\Gamma_f(r)$ is defined in
Eq.~(\ref{eq:ann_rate}). The sum runs over the kinematically allowed
annihilation final states $f$, each with a branching ratio $B_f$ and a
differential spectrum $\dd N_\e^f/\dd \ee$ that represents the number
of electrons plus positrons resulting from an annihilation event. We
use the differential spectra derived from high-statistics simulations
in \cite{2011JCAP...03..019C,2011JCAP...03..051C} to compute the
cumulative number of electrons and positrons resulting from
neutralinos annihilate indirectly into $e^+/e^-$ pairs as well as
$\mu^+/\mu^-$ pairs in the LP model. We use \ds to compute the
$e^+/e^-$ spectra from our four BM models where only a fraction of the
annihilating neutralinos is converted into electrons and positrons
(see Sect.~\ref{sect:PF} and Fig.~\ref{fig:q_DM} for further details).

@@

We focus in this section on the spectral distribution of gamma-rays
from clusters. In the LP model, DM annihilation radiation includes
final state radiation and gives rise to substantial amounts of
electrons and positrons that IC upscatter background radiation fields
to high energies. We also consider four supersymmetric DM models with
a high gamma-ray yield in the form of continuum emission and IC
induced emission. In addition to the annihilating DM, we estimate the
gamma-ray flux induced by shock-accelerated CRs. \C{Note that all
  fluxes in this section are derived within an angle corresponding to
  $\rvir$ and are convolved with a PSF of $0.1\deg$.}

@@

\begin{figure}
 \includegraphics[width=0.99\columnwidth]{figures/flux.int.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
 \caption{\it (Color online) Comparing the energy integrated flux from
   different models. We show the emission from CR induced emission
   (red solid), a leptophilic (LP) model that includes both final
   state radiation and IC upscattered CMB, dust and starlight (green
   dash-dotted), and the benchmark $\Kp$ model (BM) that includes
   continuum emission, and IC upscattered CMB, dust and starlight
   (blue dashed). The black arrow shows the spatially extended
   integrated flux upper limit set by \Fermi-LAT again indicating
   challenges for the assumptions underlying the LP model. We also
   show projected CTA point source sensitivities ($5\sigma$, 50h). The
   emission is calculated for the Fornax cluster using a point spread
   function of $0.1\deg$. The boost from Sommerfeld and substructures
   is about 530 and 890, respectively.}
 \label{fig:flux_int}
\end{figure}

>>>>

% --- journals --- %
\newcommand{\aap}{A{\&}A}
\newcommand{\aaps}{A{\&}AS}
\newcommand{\aj}{AJ}
%\newcommand{\apj}{ApJ}
\newcommand{\apjl}{ApJL}
\newcommand{\apjs}{ApJS}
\newcommand{\apss}{Ap{\&}SS}
\newcommand{\araa}{ARA{\&}A}
%\newcommand{\prd}{Phys. Rev. D}
%\newcommand{\pre}{PRE}
\newcommand{\mnras}{MNRAS}
\newcommand{\ssr}{SSRv}
%\newcommand{\nat}{Nature}
\newcommand{\physrep}{Phys.~Rep.}
\newcommand{\jcomp}{J.~Comput.~Phys.}
\newcommand{\jcap}{JCAP}
\newcommand{\Fermi}{{\em Fermi}\xspace}

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
\newcommand{\egt}{\tilde{E}_\gamma}
\newcommand{\eet}{\tilde{E}_\rmn{e}}
\newcommand{\bra}{\langle}
\newcommand{\ket}{\rangle}
\newcommand{\vst}{\vspace{-0.14mm}}
\newcommand{\vstt}{\vspace{-0.0mm}}
\newcommand{\el}{\rmn{el}}
\newcommand{\mytimes}{\hspace{-0.5mm}\times\hspace{-0.5mm}}
\newcommand{\hst}{\hspace{-1.0mm}}
\newcommand{\degs}{^\circ}
\newcommand{\colo}{(color online). }

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). Furthermore,
since most of the gamma-ray flux is expected to come from dense
subhalos within clusters which have a velocity dispersion close to the
velocity limit $v_\rmn{sat}$ where the SFE saturates, we use
$S=\B_\sfe(v_\rmn{sat})\approx 530$ for the cluster
halos.\footnote{Note that $S$ is strictly speaking the Sommerfeld
  enhanced annihilation cross section, but here we have additionally
  included the order unity constraint that ensures that the relic
  density is not over-depleted at freeze-out.} However, for figures
where we neglect the substructure boost, we adopt a velocity dependent
SFE that is normalized to fit the electron and positron excess observed at
Earth\del{observed by the PAMELA and \Fermi satellites},
\begin{equation}
\B_\sfe(\sigv) = \B_\rmn{sfe,MW} \frac{v_\rmn{MW}}{\sigv} =
70 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
\C{where the local boost factor required to explain the data is
  $\B_\rmn{sfe,MW}=300$.} \del{Here the boost factor needed to explain
  the PAMELA/\Fermi excess of electrons and positrons
  $\B_\rmn{sfe,MW}=300$,} The velocity dispersion for the MW,
$v_\rmn{MW} \approx 220\,\rmn{km\,s}^{-1}$, and the \C{mass dependent}
velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. \del{Note that, clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/fit.ds.flux.bw.eps}
 \includegraphics[width=0.49\columnwidth]{figures/fit.epflux.int.bw.eps}
 \caption{\colo Source functions for different dark
   matter (DM) models. We show the simulated data from {\sc DarkSUSY}
   \cite{ds}, and that generated by
   \protect\cite{2011JCAP...03..019C,2011JCAP...03..051C}; the solid
   lines show the fit to the data. Left panel: normalized differential
   continuum spectra for four different DM benchmark (BM) models;
   $\Ip$ model (red circles), $\Jp$ (orange squares), $\Kp$ (green
   diamonds), and $\Js$ (blue triangles). We use Eq.~(\ref{eq:bm_cont})
   to fit the continuum spectra. Right panel: number of electron and
   positron per DM annihilation above the electron energy $\ee$ for
   different DM models; $\Ip$ BM model (red circles), $\Jp$ BM model
   (orange squares), $\Kp$ BM model (green diamonds), and $\Js$ BM
   model (blue triangles), leptophilic (LP) DM annihilating indirectly
   into electrons and positrons (purple $+$) and into muons (black
   $\times$) without considering Sommerfeld boosts. We use
   Eq.~(\ref{eq:bm_elec}) and Eqs.~(\ref{eq:lp_mu}-\ref{eq:lp_ep}) to
   fit the spectra of electrons and positrons from BM and LP models,
   respectively.}
 \label{fig:q_DM}
\end{minipage}
\end{figure*}

@@

by $\qCR(\eg,r)$ (see Sec.~\ref{sect:CRs} and

@@

(see Eq.~\ref{eq:B_sfe} and Sec.~\ref{sect:LP}) and from substructure

@@

Eqs.~\ref{eq:rho_sub}-\ref{eq:xvir} and Sec.~\ref{sect:subst}). The

@@

We now derive the squared density profile for the substructures
using
\begin{eqnarray}
\rho_\rmn{sub}^2(r) &=& \frac{\dd L_\rmn{sub}}{\dd V} \frac{1}{A_{\rmn{dm}}}\,,\label{eq:rho_sub}\\
L_{200\sm} &=& A_{\rmn{dm}} \int_{\rvir} \dd r' 4\pi r'^2 \rho^2(r')\,,\label{eq:Lsm}\\
 \frac{\dd L_\rmn{sub}}{\dd V} &=& a_0\,a_1\,C(\mvir)\,x^{g(x)}\,
\left[\frac{1+a_2\log(x)}{4\pi\,r^3}\right]\,,\nonumber\\
\\
x &=& r/\rvir\,,\quad g(x) = a_2+a_1\,x^{a_2}\,.\label{eq:xvir}
\end{eqnarray}
Here, $A_{\rmn{dm}} = q_{\rmn{sm},i}(\eg,R) / \rho(R)^2$ represents
the particle physics factor, where $q_{\rmn{sm},i}(\eg,R)$ is defined
by Eq.~(\ref{eq:q_sm}). The different density profiles have some
impact on the luminosity from annihilating DM, although the details of
the density profile can be neglected compared to the dominating boost
from substructures (assuming DM to be cold). In
Fig.~\ref{fig:radial_lum} we compare the radial dependence of the
accumulative luminosity from different smooth cluster density profiles
to the boosted luminosity due to substructure for different mass
scales. We recalculate the overdensity, $\rhos$, for the shallower
density profile with $\beta=0.6$ and rescale the concentration
parameter in Eq.~(\ref{eq:cfit}) with $300/160$
\cite{2011ApJ...728L..39N} to account for the more centralized scale
radius in cluster with a shallow inner slope. The emission of this
profile with $\beta=0.6$ is about 30\% larger within $\rvir$ compared
to a cuspy NFW ($\beta=1.0$). This difference is built up within
$0.1\rvir$ (i.e. close to $r_\s$). Hence the slope of the central part
of a cluster has little influence for the DM luminosity within $\rvir$
as long as the degeneracy in the lensing mass measurements with $r_\s$
(which decreases for decreasing inner slope) has been taken into
account. The emission from an Einasto density profile is about 50\%
larger than the cuspy NFW profile in the periphery of the cluster,
where the difference is mainly built up at a few percent of
$\rvir$. The increase in luminosity due to substructures is negligible
in the center of halos, but integrated out to $\rvir$ the boost
relative to the smooth emission profile amounts to approximately 20
for dwarf galaxies, $200$ for galaxies, and $10^3$ for galaxy
clusters. \C{We stress that these boost factors are only realized in
  the region around the virial radius of each respective halo which
  is mostly tidally stripped for dwarfs in the MW. Hence a more
  realistic boost from substructures is probably much smaller in
  satellite dwarf galaxies. In addition,} these boost factors are only
realized for direct annihilation emission (continuum emission or final
state radiation) or IC scattering off homogeneous seed photon fields
(CMB). For IC scattering of SD photons, the overlap of final state
leptons and SD photons is smaller which causes the substructure boost
over the smooth emission to be reduced by roughly two orders of
magnitudes. It should also be noted that the substructure boosted
fluxes from clusters are much more extended than for dwarf galaxies,
hence more difficult to detect with Cherenkov telescopes. In fact, it
was shown in Sanchez-Conde et al. \cite{2011arXiv1104.3530S} that
inside $0.1\degs$, where the sensitivity of Cherenkov telescopes is
maximized, the expected DM flux including substructures from the
brightest dwarf galaxy is about an order magnitude higher than the
brightest cluster. However, that work assumed a substructure boost of
about 50 in the most massive cluster halos which is more than an order
magnitude smaller than what we use in this work, hence in projection
we expect this difference to be a factor few smaller.

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/dens.prof.bw.eps}
 \caption{\colo Radial dependence of DM annihilation
   luminosity of smooth halo and substructures. The solid lines show
   the accumulative smooth luminosity from a cluster with the mass
   $\mvir=10^{15}\,\msun$ for three different density profiles; an
   Einasto profile with $\alpha=0.17$ (light blue), a cuspy NFW
   profile with $\beta=1.0$ (thick dark blue), and a core NFW profile
   with $\beta=0.6$ (thin purple). The dashed lines show the
   accumulative luminosity from substructures for three different mass
   scales; an $\mvir=10^{15}\,\msun$ galaxy cluster (thick red), an
   $\mvir=10^{12}\,\msun$ galaxy (orange), and an
   $\mvir=10^{8}\,\msun$ dwarf galaxy (thin green). All luminosities
   have been normalized with the luminosity within $\rvir$ from a
   cuspy NFW profile. We have assumed the standard value for the
   limiting substructure mass of $M_\rmn{lim}=10^{-6}\,\msun$. Note
   the large expected boost from substructures in clusters
   ($\sim1000$), and the relative small boost in dwarf galaxies
   ($\sim20$).}
 \label{fig:radial_lum}
\end{figure}

In Fig.~\ref{fig:radial_emis} we show the radial regions that dominate
the DM annihilation luminosity, in particular we show the differential
contribution to the DM luminosity per logarithmic interval in radius
for three different mass scales. Solving for the maximum of this
curve, $\dd^2L_{200\sm}/\dd (\log x)^2=0$ in combination with
Eq.~(\ref{eq:Lsm}), we find that the luminosity from the smooth NFW
profile peaks at $r \simeq \rvir/3c$. Despite the cuspy nature of the
density profile, the luminosity is not dominated by the central region
but by the transition region where the profile steepens because of the
larger volume available there. For large clusters with typical
concentrations of $c = \rvir/r_\s \simeq 4$, the luminosity from the
smooth profile is focused to the regime around 10\% of $\rvir$. In
contrast, emission from substructures is mainly contributed by the
outer parts of DM halos. As shown in Fig.~\ref{fig:radial_emis}, the
product of annihilation emissivity and emission volume increases
towards $\rvir$ and only starts to drop outside this radius. Note
that, even though most of the substructure mass density has been
erased in the central regions of DM halos, a cluster in projection has
a significant enhancement due to substructures at a radius of just a
few percent of $\rvir$.
\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/emissiv.sub.bw.eps}
  \caption{\colo Plot that shows where most of the DM
    annihilation luminosity originates. We show the differential
    contribution to the DM annihilation luminosity per logarithmic
    interval in radius which corresponds to $x^3\rho^2/\rho^2_{200}$
    where $\rho_{200}$ is the density at $\rvir$. The solid lines
    represent cuspy $\beta=1.0$ NFW density profiles for three
    different mass scales; an $\mvir=10^{15}\,\msun$ galaxy cluster
    (thick red), an $\mvir=10^{12}\,\msun$ galaxy (orange), and an
    $\mvir=10^{8}\,\msun$ dwarf galaxy (thin green). The dashed lines
    show the contribution from substructures for the same three mass
    scales. All luminosities have been normalized with the luminosity
    within $\rvir$ from a cuspy NFW profile. We have assumed the
    standard value for the limiting substructure mass of
    $M_\rmn{lim}=10^{-6}\,\msun$. For the smooth profile, the majority
    of the flux is delivered by a region around $r_\s/3$ as indicated
    by the maximum value of the curves. In contrast, for substructures
    the emission is dominated by regions around $r_{200}$.}
  \label{fig:radial_emis}
\end{figure}

@@

The electrons injected from annihilating DM also suffer from diffusive
and radiative losses. Hence we have to calculate the equilibrium
spectrum of the electrons plus positrons denoted by $\dd n_\e/\dd \ee$
in Eq.~(\ref{eq:ICemiss}). We derive this stationary solution using
the cosmic ray transport equation (neglecting convection and
re-acceleration effects):
\begin{eqnarray}
\label{eq:CRdiffusion}
\frac{\partial}{\partial t}\left(\frac{\dd n_\e}{\dd \ee}\right) =
&\nabla& \left[D\left(\ee,\bx\right)\nabla\frac{\dd n_\e}{\dd \ee}\right] +
\frac{\partial}{\partial \ee}
\left[b\left(\ee,\bx\right) \frac{\dd n_\e}{\dd \ee}\right]
 \nonumber \\
&+& q_\e(\ee,\bx)\,,
\end{eqnarray}
where $D(\ee,\bx)$ denotes the diffusion coefficient and $b(\ee,\bx)$
the energy loss term. The source function $q_\e(\ee,\bx)$ yields the
number of electrons and positrons produced per unit time, energy and
volume element at the position $\bx$:
\begin{equation}
q_\e(\ee,r)=\sum_f\frac{\dd N_\e^f}{\dd \ee}(\ee) B_f \Gamma_f(r) \,,
\end{equation}
where the annihilation rate density $\Gamma_f(r)$ is defined in
Eq.~(\ref{eq:ann_rate}). The sum runs over the kinematically allowed
annihilation final states $f$, each with a branching ratio $B_f$ and a
differential spectrum $\dd N_\e^f/\dd \ee$ that represents the number
of electrons plus positrons resulting from an annihilation event. We
use the differential spectra derived from high-statistics simulations
in \cite{2011JCAP...03..019C,2011JCAP...03..051C} to compute the
cumulative number of electrons and positrons resulting from
neutralinos annihilate indirectly into $e^+/e^-$ pairs as well as
$\mu^+/\mu^-$ pairs in the LP model. We use \ds to compute the
$e^+/e^-$ spectra from our four BM models where only a fraction of the
annihilating neutralinos is converted into electrons and positrons
(see Sec.~\ref{sect:PF} and Fig.~\ref{fig:q_DM} for further details).

@@

We focus in this section on the spectral distribution of gamma-rays
from clusters. In the LP model, DM annihilation radiation includes
final state radiation and gives rise to substantial amounts of
electrons and positrons that IC upscatter background radiation fields
to high energies. We also consider four supersymmetric DM models with
a high gamma-ray yield in the form of continuum emission and IC
induced emission. In addition to the annihilating DM, we estimate the
gamma-ray flux induced by shock-accelerated CRs. \C{Note that all
  fluxes in this section are derived within an angle corresponding to
  $\rvir$ and are convolved with a PSF of $0.1\degs$.}

@@

\begin{figure}
 \includegraphics[width=0.99\columnwidth]{figures/flux.int.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
 \caption{\colo Comparing the energy integrated flux from
   different models. We show the emission from CR induced emission
   (red solid), a leptophilic (LP) model that includes both final
   state radiation and IC upscattered CMB, dust and starlight (green
   dash-dotted), and the benchmark $\Kp$ model (BM) that includes
   continuum emission, and IC upscattered CMB, dust and starlight
   (blue dashed). The black arrow shows the spatially extended
   integrated flux upper limit set by \Fermi-LAT again indicating
   challenges for the assumptions underlying the LP model. We also
   show projected CTA point source sensitivities ($5\sigma$, 50h). The
   emission is calculated for the Fornax cluster using a point spread
   function of $0.1\degs$. The boost from Sommerfeld and substructures
   is about 530 and 890, respectively.}
 \label{fig:flux_int}
\end{figure}
